Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-07-27 paper/manuscript.tex

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020automated}.
\end{itemize}

@@

Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 
 This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020automated}.

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of benchmark datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well.
For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, but an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}. Here the errors refer to the difference in $\Delta G$ after fitting to the mean experimental binding energy. This prompts us to consider another important aspect. It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges. (See also the section recommendations for best practices of reporting data Sec.~\ref{sec:plot_data}) In general, the computed edge $\Delta \Delta G$ RMSEs significantly underestimates the error of all edge RMSEs: \url{https://github.com/jchodera/jacs-dataset-analysis}. Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each particular application case.

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
Finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol)~\cite{rizzi2019sampl6}.

For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. A good practice is to perform two or three runs of the same perturbation to assess repeatability with a given protocol. The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. Where this is practical, a more stringent test is to use different input coordinates for each repeat run.

Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently. Greater variability is expected due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported, even after fixes to the underlying simulation packages~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions.

@@

To ensure modeling matches experiment, we also need to accurately prepare and  simulate the same system -- which requires understanding what protein construct is used in the bioassay. For instance, does the X-ray structure that is to be used for the calculations match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer etc)~\cite{perez-benito2018predicting}? Also, were certain co-factors or partner proteins required in the bioassay? 
%
\subsection{Is my binding mode accurate?}
As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, usually from a high quality X-ray crystal structure. If more than one structure is available, the modeler should pay attention to choose the most suitable. The quality of the structure can be a concern, and the reader is referred to work of Warren et al. for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}.

It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there is evidence of this in any alternative X-ray structures of the same protein. Often, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations. This can provide a practical limit on the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.

@@

Finally, the role of water in ligand binding is not always well understood and it can be crucial to capture the changes in binding site solvation during ligand binding. Can crystallographic waters be retained? Do they clash with some of the larger ligands used in the alchemical perturbation? If so, how will the solvent be incorporated for smaller ligands that require the water -- consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}. Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Before launching large numbers of alchemical free energy calculations it is recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations. Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for alchemical calculations.

Running free energy of binding calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations. The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes. The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs. Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace, kuhn2020automated}.

@@

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.

@@

\subsection{Free energies of binding}
\label{subsec:binding}
In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice because simulations are of finite duration (typically 1-100 ns per state at present) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. 
The following questions may be relevant before diving into the simulation setup.

@@

Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see~\ref{sec:simulation_protocol_choice}). For relative calculations, some care has to be taken selecting binding poses for these ligands. Generally a common assumption for a congeneric series is that the binding mode is conserved. Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation without steric or electrostatic mismatch with the receptor. Checking the X-ray structure versus the experimental electron densities is important, as the position of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density, especially in cases of missing density. For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and, if a boat conformation is present in the structure, it may be worth inspecting the density to ensure it adequately supports this choice.

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful. For examples, see HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.

@@

\paragraph{How to align congeneric series}
Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms. The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement. This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series. Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection, but there is a risk that the complex relaxes to an alternative metastable state.

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}). In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations. In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.

If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites. This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{evoli2016multiple}. The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind. In this case a docking program can generate initial structures. Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, kuhn2020automated}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What simulation protocol should I choose   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What simulation protocol should I choose?}
\label{sec:simulation_protocol_choice}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.

@@

The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g. consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.
In this case the common substructure is the benzene ring, though may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. On this new carbon you will also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group after the transformation. Bond parameters as well as partial charges are adjusted accordingly. 
Thus, in a single topology calculation, atoms may change their type ensuring only relatively few dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. 
In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. 
\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. \textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences. Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:fig_topology}
\end{figure} 
Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time (turning one molecule off while turning the other on)~\cite{jiang2019computing}.
At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2015accurate,wang2019protein} and in BioSimSpace~\cite{hedges2019biosimspace} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.
To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead,mobley2012perspective}.
%
\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed.}
    \label{fig:fig_mcss}
\end{figure}

@@

MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.

>>>>

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
\end{itemize}

@@

Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 
 This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020assessment}.

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well.
For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, but an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}. Here the errors refer to the difference in $\Delta G$ after doing a linear regression fit to observed experimental binding free energies. This prompts us to consider another important aspect. It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges. (See also the section recommendations for best practices of reporting data Sec.~\ref{sec:plot_data}) In general, the computed edge $\Delta \Delta G$ RMSEs significantly underestimates the error of all edge RMSEs: \url{https://github.com/jchodera/jacs-dataset-analysis}. Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each particular application case.

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
Finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol)~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. A good practice is to perform two or three runs of the same perturbation to assess repeatability with a given protocol. The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. Where this is practical, a more stringent test is to use different input coordinates for each repeat run.

Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently. Greater variability is expected due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. Further work is needed to ensure reproducibility of alchemical free energy calculations across different software implementations to guarantee that force-field development efforts lead to transferable potential energy functions.

@@

To ensure modeling matches experiment, we also need to accurately prepare and  simulate the same system -- which requires understanding what protein construct is used in the bioassay. For instance, does the X-ray structure that is to be used for the calculations match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer etc)~\cite{perez-benito2018predicting}? Also, were certain co-factors or partner proteins required in the bioassay?

\subsection{Is my binding mode accurate?}
As also mentioned, good performance of alchemical calculations requires an accurate representation of the ligand binding mode, usually from a high quality X-ray crystal structure. If more than one structure is available, the modeler should pay attention to choose the most suitable. The quality of the structure can be a concern, and the reader is referred to work of Warren et al. for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}.

It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there is evidence of this in any alternative X-ray structures of the same protein. Often, only one protein and water configuration is used for a series of alchemical calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations. This can provide a practical limit on the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.

@@

Finally, the role of water in ligand binding is not always well understood and it can be crucial to capture the changes in binding site solvation during ligand binding. Can crystallographic waters be retained? Do they clash with some of the larger ligands used in the alchemical perturbation? See Sec.~\ref{subsec:binding}  for different strategies that can be applied to dealing with waters. Generally, before launching large numbers of alchemical free energy calculations it is always recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations. Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for alchemical calculations.

Running free energy of binding calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations. The commercial implementation such as FEP+, OpenEye Tools, or Flare allow for a fast setup and deployment to GPU hardware in minutes, but will have limited ability to customize calculations~\cite{wang2015accurate,kuhn2020assessment}.  Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace, kuhn2020assessment}.

@@

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical calculations.

@@

\subsection{Free energies of binding}
\label{subsec:binding}
In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice because simulations are of finite duration (typically 1-100 ns per state at present) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that significantly exceed the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. 
The following questions may be relevant before diving into the simulation setup.

@@

Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see~\ref{sec:simulation_protocol_choice}). For relative calculations, some care has to be taken selecting binding poses for these ligands. Generally, a common assumption for a congeneric series is that the binding mode is conserved. Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation without steric or electrostatic mismatch with the receptor. Checking the X-ray structure versus the experimental electron densities is important, as the position of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density, especially in cases of missing density. For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and, if a boat conformation is present in the structure, it may be worth inspecting the density to ensure it adequately supports this choice.

@@

\paragraph{Conserved binding site waters can play an important role in binding free energies}
Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales. This happens typically in buried binding sites. Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful, i.e. consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}. Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}. Well known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.

@@

\paragraph{How to align congeneric series}
Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure (MCSS) algorithms. The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement. This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series. Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection, but there is a risk that the complex relaxes to an alternative metastable state.

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}). In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations. In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.

If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites. This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{evoli2016multiple}. The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind. In this case a docking program can generate initial structures. Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, kuhn2020assessment}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What simulation protocol should I choose   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What simulation protocol should I choose?}
\label{sec:simulation_protocol_choice}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.

@@

The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g. consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.
In this case the common substructure is the benzene ring, though may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. On this new carbon you will also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group after the transformation. Bond parameters as well as partial charges are adjusted accordingly. 
Thus, in a single topology calculation, atoms may change their type ensuring only relatively few dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. 
In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. 
\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. \textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences. Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:fig_topology}
\end{figure} 
Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time (turning one molecule off while turning the other on)~\cite{jiang2019computing}.
Various software implementations currently mostly rely on single or dual topology approaches; for example AMBER TI uses a dual topology approach, while Gromacs and BioSimSpace use single topology approaches. Please make sure to check what approach is used with your software package of choice. 
To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead, mobley2012perspective}.

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed.}
    \label{fig:fig_mcss}
\end{figure}

@@

MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}. The MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
